Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-13 talk/icooolps2011/paper.tex
michael's notes

Meta-tracing JITs can be applied to a variety of different
languages without explicitly encoding language semantics into the compiler. So
far, they lacked a way to feed back runtime information into the
compiler, which restricted their performance. In this paper we describe the
flexible mechanisms in PyPy's meta-tracing JIT that can be used to control runtime feedback in language-specific ways. These mechanisms are flexible
enough to implement classical VM techniques such as maps and polymorphic inline
caches.

@@

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have rather complex core object
semantics. For them, even implementing just an interpreter is already a complex
task. Implementing them efficiently with a just-in-time compiler (JIT) is
extremely challenging, because of their many corner-cases.

@@

Concretely these hints are used to control how the optimizer of the
tracing JIT can improve the traces of the object model. In particular the hints
influence the constant folding
optimization. The first hint makes it possible to turn arbitrary
variables in the trace into constant by feeding back runtime values. The
second hint allows the definition of additional foldable operations.

Together these two hints can be used to express many classic implementation
techniques used for object models of dynamic languages, such as maps and
polymorphic inline caches.

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace, that
 means the feedback of runtime information into compilation.
 \item A way to define new pure operations which the constant folding
 optimization then recognizes.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
\end{itemize}

@@

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are therefore linear list of operations, which are optimized and then
get turned into machine code. This recording automatically inlines functions:
when a function call is encountered the operations of the called functions are
simply put into the trace of the caller too. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{andreas_gal_incremental_2006}.

@@

One disadvantage of (tracing) JITs which makes them not directly applicable to
PyPy is that they need to encode the language semantics of the language they are
tracing. Since PyPy wants to be a
general framework, we want to reuse our tracer for different languages.
Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

While the operations in a trace are those of the interpreter, the loops that are
traced by the tracer are the loops in the
user program. To achieve this the tracer stops tracing after one iteration of
the loop in the user function that is being considered. At this point, it probably
traced many iterations of the interpreter main loop.

@@

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other advanced features. The model has classes, which contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' methods are searched.

@@

To implement this object model, we could use the RPython code in
Figure~\ref{fig:interpreter-slow} as part of the interpreter source code.
In this straightforward implementation the methods and attributes are just
stored in dictionaries (hash maps) on the classes and instances, respectively.
While this object model is very
simple it already contains most hard parts of Python's object model. Both
instances and classes can have arbitrary fields, and they are changeable at
any time.  Moreover, instances can change their class after they have been
created.

When using this object model in
an interpreter, a huge amount of time will be spent doing lookups in these
dictionaries.
Let's assume we trace through code that sums three attributes, such as:
\anto{I still think it's a bit weird to call them ``methods'' and then use
  them as attributes in the example}

@@

The simplest example of constants are literal values.  For example, if in the
RPython source code we have a line like \texttt{y = x + 1}, the second operand will
be a constant in the trace.

However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code:

@@

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend not to vary often. Therefore it is possible to
promote the types. Section~\ref{sec:} will present a complete example of how
this works.

@@

In the previous section we saw a way to turn arbitrary variables into constants. All
pure operations on these constants can be constant-folded. This works great for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not pure and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

@@

>>>>

Meta-tracing JIT compilers can be applied to a variety of different
languages without explicitly encoding language semantics into the compiler. So
far, they lacked a way to feed back runtime information into the
compiler, which restricted their performance. In this paper we describe the
flexible mechanisms in PyPy's meta-tracing JIT that can be used to control runtime feedback in language-specific ways. These mechanisms are flexible
enough to implement classical VM techniques such as maps and polymorphic inline
caches.

\cfbolz{XXX tracing is runtime feed back too, clarify}

@@

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have a rather complex core object
semantics. For them, even implementing just an interpreter is already a complex
task. Implementing these languages efficiently with a just-in-time compiler (JIT) is
extremely challenging, because of their many corner-cases.

@@

\cfbolz{XXX kill the next paragraph? the info is repeated in the list below}
Concretely these hints are used to control how the optimizer of the
tracing JIT can improve the traces of the object model. In particular the hints
influence the constant folding
optimization. The first hint makes it possible to turn arbitrary
variables in the trace into constant by feeding back runtime values. The
second hint allows the definition of additional foldable operations.

Together these hints can be used to express many classic implementation
techniques used for object models of dynamic languages, such as maps and
polymorphic inline caches.

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace by
 feeding back runtime information into compilation.
 \item A way to annotate operations as pure which the constant folding
 optimization then recognizes.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
\end{itemize}

@@

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are linear lists of operations, which are optimized and then
get turned into machine code. This recording automatically inlines functions:
when a function call is encountered the operations of the called functions are
simply put into the trace of the caller too. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{andreas_gal_incremental_2006}.

@@

Since PyPy wants to be a general framework, we want to reuse our tracer for
different languages, which makes classical tracers inapplicable, because they
encode language semantics.
Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

While the operations in a trace are those of the interpreter, the loops that are
traced by the tracer are the loops in the
user program. To achieve this the tracer stops tracing after one iteration of
the loop in the user function that is being considered; at this point, it probably
traced many iterations of the interpreter main loop.

@@

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other advanced features. In the model classes contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' methods are searched.

@@

To implement this object model, we use the RPython code in
Figure~\ref{fig:interpreter-slow} as part of the interpreter source code.
In this straightforward implementation the methods and attributes are just
stored in dictionaries (hash maps) on the classes and instances, respectively.
While this object model is very
simple it already contains most hard parts of Python's object model. Both
instances and classes can have arbitrary fields, and they are changeable at
any time.  Moreover, instances can change their class after they have been
created.

When using this object model in
an interpreter, a large amount of time will be spent doing lookups in these
dictionaries.
Let's assume we trace through code that sums three attributes, such as:
\anto{I still think it's a bit weird to call them ``methods'' and then use
  them as attributes in the example}

@@

The simplest example of constants are literal values, such as \texttt{1}.
However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code:

@@

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend not to vary often. Therefore it is possible to
promote the types. Section~\ref{sec:fastobjmodel} will present a complete example of how
this works.

@@

In the previous section we saw a way to turn arbitrary variables into constants. All
pure operations on these constants can be constant-folded. This works well for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not pure and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

@@

\cfbolz{XXX mention a possible debug mode for findings bugs in this, or too many values in promotion. stress more that promote is safe}
